\documentclass[11pt]{article}
\usepackage[margin=0.9in]{geometry}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage[hidelinks]{hyperref}
\usepackage{xurl}
\usepackage{graphicx}
\usepackage{enumitem}
\usepackage{amsmath}
\usepackage{float}
\usepackage{tikz}
\usetikzlibrary{arrows.meta, positioning, shapes.geometric}
\setlist[itemize]{topsep=0.25em, itemsep=0.25em}
\setlist[enumerate]{topsep=0.25em, itemsep=0.25em}

\newcommand{\projectname}{Quantyze}
\newcommand{\projecttitle}{Trees, Trades, and Books}
\newcommand{\projectsubtitle}{Simulating a Limit Order Book with Tree-Based Data Structures and an ML Agent}
\newcommand{\projectdate}{March 31, 2026}

\begin{document}
\begin{center}
{\Huge \projectname: \projecttitle\par}
\vspace{0.8em}
\noindent\rule{0.82\textwidth}{0.6pt}\par
\vspace{0.9em}
{\large\itshape\parbox{0.9\textwidth}{\centering \projectsubtitle}\par}
\vspace{1.4em}
\begin{minipage}[t]{0.31\textwidth}
\centering
\large Cade McNelly
\end{minipage}
\hfill
\begin{minipage}[t]{0.31\textwidth}
\centering
\large Nicolas Miranda Cantanhede
\end{minipage}
\hfill
\begin{minipage}[t]{0.31\textwidth}
\centering
\large Sahand Samadirand
\end{minipage}

\vspace{1.2em}
{\large \projectdate\par}
\end{center}
\vspace{1.2em}

\section*{1. Introduction}
As a group of CS students interested in finance, we have spent a lot of time
looking at stock charts. It does not take long to notice a major disconnect
between the price shown on the screen and the mechanics operating underneath
it. A market interface may display a stock at a certain level, but that does
not mean an order can be filled there instantly or in unlimited size. In
practice, fills depend on liquidity, queue position, and the slippage created
when an order moves through the book.

That gap between a quoted price and an executable price is one of the main
ideas that motivated this project. It is also a very natural place to apply
CSC111 ideas such as tree-based data structures, simulation, and algorithmic
state updates. Quantyze studies that gap by simulating a limit order book in Python. We represent bids and asks as separate binary search trees of
price levels, maintain FIFO priority within each level, replay event streams
through a matching engine, and compute summary statistics about fills,
cancellations, spread, midpoint, and slippage. On top of the simulator, we
added a small machine learning component that observes book-state
features and classifies market actions in the short term.

\textbf{Our project goal is to build a tree-based market simulator that can
replay realistic order flow, expose the mechanics of price-time-priority
matching, and train a small neural network to classify short term market actions
as buy, sell, or hold from order-book state.}

The project kept the same core direction as our proposal, but the final version
organizes the submission around a browser display launched from
\texttt{main.py}, with a terminal-menu fallback for the same simulator and
training logic. That let us present the project in a PyCharm-friendly way without
changing the main computational challenge: representing and updating a live
order book efficiently with trees while connecting the simulator to a
meaningful data and model-training pipeline. In the language of our original
proposal, we still see the project as an attempt to close the gap between what
a trader sees on a chart and what actually has to happen for an order to be
filled.

\section*{2. Datasets}
One thing that became much clearer while building the final project is that not
all data in Quantyze plays the same role. Quantyze uses three distinct
\textbf{data-source roles}, and separating them clearly makes the rest of the
system much easier to follow:
\begin{itemize}
    \item \textbf{synthetic scenarios} for simulator-first replay demos
    \item \textbf{packaged internal CSVs} for fast and reproducible classifier
    retraining
    \item \textbf{paired raw LOBSTER files} for the real-data preprocessing and
    feature-building path
\end{itemize}

The submission also includes a packaged dataset zip, so the TA does not need to
download datasets or baseline model artifacts externally before running the
project. That was important to us because we wanted the grading path to be
packaged, local, and easy to repeat.

\subsection*{2.1 Synthetic Event Data}
We generated our own synthetic event streams for testing and demonstration.
These are created inside \texttt{data\_loader.py} and include three scenarios:
\begin{itemize}
    \item \texttt{balanced}
    \item \texttt{low\_liquidity}
    \item \texttt{high\_volatility}
\end{itemize}

These synthetic datasets use the same canonical event format as the rest of the
project: \texttt{timestamp}, \texttt{order\_id}, \texttt{side},
\texttt{order\_type}, \texttt{price}, and \texttt{quantity}. Their role is
\textbf{simulation only}. They are the fastest way to see fills, cancellations,
spread changes, and book updates without
waiting for a large dataset to load.

\subsection*{2.2 LOBSTER Sample Data}
For the real-data path of the machine learning portion of the project, we used
LOBSTER sample order-book data from the official LOBSTER sample page.\footnote{\url{https://data.lobsterdata.com/info/DataSamples.php}}
Each LOBSTER sample comes as a \textbf{pair of files}:
\begin{itemize}
    \item a message file containing timestamped market events
    \item an order-book file containing depth snapshots
\end{itemize}

Our code supports raw LOBSTER message and order-book CSVs. The training flow is
pointed at the \textbf{message file}, and the loader infers the paired
order-book file automatically from the filename. From the message file, we use
normalized information corresponding to timestamp, event type, order id, size,
price, and direction. The loader also recognizes message types that should be
preserved as annotations rather than replayed directly.

From the paired order-book file, our current feature pipeline uses only the top
two bid levels and top two ask levels, even when a file contains deeper book
data. In practice, the model ultimately trains on level-1 and level-2
price/size information rather than the full book.

During development, we used public LOBSTER sample files such as AAPL, AMZN, and
GOOG to validate the real-data pipeline. In the final submission, we ship one
official Apple 5-level sample pair inside the dataset zip using the clearer
extracted filenames:
\begin{itemize}
    \item \texttt{aapl\_lobster\_2012-06-21\_message\_5level\_sample.csv}
    \item \texttt{aapl\_lobster\_2012-06-21\_orderbook\_5level\_sample.csv}
\end{itemize}

This packaged LOBSTER option is \textbf{optional}. It demonstrates the
real-data pipeline, but it is much larger and slower to load than the short
internal packaged dataset.

\subsection*{2.3 Submission Dataset Package}
For the final submission, the dataset zip is \texttt{quantyze\_datasets.zip}. It
contains:
\begin{itemize}
    \item \texttt{sample\_internal.csv}
    \item \texttt{huge\_internal.csv}
    \item \texttt{aapl\_lobster\_2012-06-21\_message\_5level\_sample.csv}
    \item \texttt{aapl\_lobster\_2012-06-21\_orderbook\_5level\_sample.csv}
    \item \texttt{instructions.txt}
    \item \texttt{model.pt}
    \item \texttt{training\_metrics.json}
\end{itemize}

After extraction, these files are expected to sit beside \texttt{main.py}. Both
the browser UI and the terminal menu look for the packaged datasets and
shipped baseline artifacts there.

The intended roles of these packaged files are:
\begin{itemize}
    \item \texttt{sample\_internal.csv}: the \textbf{quick} retraining check
    \item \texttt{huge\_internal.csv}: the \textbf{large internal} retraining
    demo, which takes longer to load and train
    \item \texttt{aapl\_lobster\_2012-06-21\_message\_5level\_sample.csv}: the
    \textbf{large raw LOBSTER message-file} demo
    \item \texttt{aapl\_lobster\_2012-06-21\_orderbook\_5level\_sample.csv}:
    the paired order-book file required by the raw LOBSTER path
\end{itemize}

The package also includes saved metrics and a baseline checkpoint, so the TA
can use the model path without retraining first.

The packaged datasets and baseline artifacts are local. The one browser-side
runtime dependency that is still external is the \texttt{Chart.js} CDN used by
the \texttt{Charts} tab, so viewing those browser charts assumes a normal
internet connection even though the datasets and checkpoints are bundled with
the submission.

\section*{3. Computational Overview}
\subsection*{3.1 Core Data Representation}
The central data structure in Quantyze is the limit order book. We represent the
book with two binary search trees:
\begin{itemize}
    \item one tree for bids
    \item one tree for asks
\end{itemize}

This logic is implemented across \texttt{book\_tree.py},
\texttt{price\_level.py}, and \texttt{order\_book.py}. Each node in a tree is a
\texttt{PriceLevel} object. A price level stores:
\begin{itemize}
    \item a price
    \item aggregate resting volume at that price
    \item a FIFO queue of resting orders
\end{itemize}

The trees let us preserve sorted price structure while still supporting
efficient best-bid and best-ask access. The FIFO queue inside each price level
enforces time priority among orders resting at the same price. The queue is
implemented as a doubly-linked list with a parallel \texttt{order\_id}-to-node
dictionary, which gives O(1) cancellation by order id in addition to O(1)
enqueue and dequeue. This combination directly models price-time-priority
matching, which was the main algorithmic goal of our project.\footnote{Gould
et al.\ (2013) provide a comprehensive treatment of price-time priority and
order-book microstructure dynamics:
\url{https://arxiv.org/abs/1012.0349}}

\begin{figure}[H]
\centering
\begin{tikzpicture}[
    node distance=0.55cm and 0.9cm,
    every node/.style={font=\small},
    level/.style={draw, rectangle, rounded corners=2pt,
                  minimum width=1.7cm, minimum height=0.55cm,
                  text centered},
    bid/.style={level, fill=blue!15},
    ask/.style={level, fill=red!12},
    root/.style={level, fill=gray!20, minimum width=2.2cm},
    queue/.style={draw, rectangle, minimum width=0.9cm,
                  minimum height=0.42cm, font=\scriptsize, fill=yellow!25},
    arr/.style={-{Stealth[length=4pt]}, thin},
]
\node[root] (ob) {OrderBook};
\node[bid, below left=0.9cm and 2.0cm of ob]  (bid_root) {100.0 bid};
\node[bid, below left=0.5cm and 0.7cm of bid_root]  (bid_l) {99.8 bid};
\node[bid, below right=0.5cm and 0.7cm of bid_root] (bid_r) {99.5 bid};
\node[ask, below right=0.9cm and 2.0cm of ob] (ask_root) {100.2 ask};
\node[ask, below left=0.5cm and 0.7cm of ask_root]  (ask_l) {100.4 ask};
\node[ask, below right=0.5cm and 0.7cm of ask_root] (ask_r) {100.6 ask};
\draw[arr] (ob.south west) -- (bid_root.north);
\draw[arr] (ob.south east) -- (ask_root.north);
\draw[arr] (bid_root.south west) -- (bid_l.north);
\draw[arr] (bid_root.south east) -- (bid_r.north);
\draw[arr] (ask_root.south west) -- (ask_l.north);
\draw[arr] (ask_root.south east) -- (ask_r.north);
\node[above=0.05cm of bid_root, font=\scriptsize\itshape, text=blue!60] {best bid = max};
\node[above=0.05cm of ask_root, font=\scriptsize\itshape, text=red!60]  {best ask = min};
\end{tikzpicture}
\caption{The \texttt{OrderBook} holds two \texttt{BookTree} instances. The bid
tree's maximum-price node is the best bid; the ask tree's minimum-price node is
the best ask. Each \texttt{PriceLevel} node holds a FIFO doubly-linked-list
queue of resting orders that enforces
time priority.}
\label{fig:order-book}
\end{figure}

\subsection*{3.2 Event Replay and Matching}
All incoming data is converted into a common \texttt{Event} representation in
\texttt{orders.py}. The \texttt{DataLoader} in \texttt{data\_loader.py} either:
\begin{itemize}
    \item loads a dataset from internal CSV or raw LOBSTER files, or
    \item generates a synthetic scenario
\end{itemize}

Those events are then replayed through \texttt{EventStream} in
\texttt{event\_stream.py}, which emits them one at a time into the
\texttt{MatchingEngine} in \texttt{matching\_engine.py}. The matching engine
handles limit orders, market orders, and cancellations. It updates the
\texttt{OrderBook}, performs partial and full fills, removes empty price levels
when necessary, and records execution information used for later metrics.

At the end of a normal run, the simulation runtime reports:
\begin{itemize}
    \item total filled quantity
    \item fill count
    \item cancel count
    \item average slippage
    \item final spread
    \item final midpoint
    \item agent overlay mark-to-market P\&L if a trained checkpoint is active
\end{itemize}

The same simulation and training state is surfaced in two ways:
\begin{itemize}
    \item the browser UI launched by \texttt{main.py}
    \item the optional terminal menu opened through
    \texttt{main(run\_ui=False)}
\end{itemize}

The saved artifacts that support those workflows include:
\begin{itemize}
    \item \texttt{model.pt}
    \item \texttt{training\_metrics.json}
    \item \texttt{latest\_model.pt}
    \item \texttt{latest\_training\_metrics.json}
    \item \texttt{latest\_training\_data.csv}
    \item \texttt{active\_model.json}
    \item \texttt{log.json}
\end{itemize}

The packaged baseline artifacts are \texttt{model.pt} and
\texttt{training\_metrics.json}. The remaining files are created or updated as
the TA simulates, retrains, changes overlay mode, or inspects the latest
execution log.

\subsection*{3.3 Data Processing and Training Pipeline}

\begin{figure}[H]
\centering
\begin{tikzpicture}[
    node distance=0.45cm and 0.6cm,
    box/.style={draw, rectangle, rounded corners=2pt,
                minimum width=2.3cm, minimum height=0.58cm,
                text centered, font=\small, fill=gray!10},
    arr/.style={-{Stealth[length=4pt]}, thin},
    lab/.style={font=\scriptsize\itshape, text=gray!80},
]
\node[box] (events)  {Event Stream};
\node[box, right=1.1cm of events]  (engine) {Matching Engine};
\node[box, right=1.1cm of engine]  (book)   {OrderBook};
\node[box, below=0.8cm of book]    (feat)   {12 base features};
\node[box, below=0.5cm of feat]    (aug)    {+4 delta features};
\node[box, below=0.5cm of aug]     (label)  {horizon label};
\node[box, left=1.1cm of label]    (net)    {OrderBookNet};
\node[box, left=1.1cm of net]      (ckpt)   {checkpoint (.pt)};
\draw[arr] (events)  -- node[above, lab]{replay} (engine);
\draw[arr] (engine)  -- node[above, lab]{updates} (book);
\draw[arr] (book)    -- node[right, lab]{\texttt{depth\_snapshot()}} (feat);
\draw[arr] (feat)    -- (aug);
\draw[arr] (aug)     -- (label);
\draw[arr] (label)   -- node[below, lab]{loss} (net);
\draw[arr] (net)     -- node[below, lab]{\texttt{save()}} (ckpt);
\end{tikzpicture}
\caption{This is how the training data flows: events are replayed through the matching engine to
keep the \texttt{OrderBook} live. At each step, a 16-D feature vector (12
instantaneous book-state features plus 4 one-step deltas) and a horizon-based
label are extracted. \texttt{OrderBookNet} is trained with class-weighted
cross-entropy and the best-validation-loss state is saved as a checkpoint.}
\label{fig:pipeline}
\end{figure}

The machine learning pipeline is built mainly in \texttt{data\_loader.py},
\texttt{neural\_net.py}, and \texttt{training.py}.

The feature pipeline uses 16 features per example. The first 12 are instantaneous
book-state features extracted by
\texttt{DataLoader.\_feature\_vector\_from\_levels}:
\begin{enumerate}
    \item best bid price
    \item best bid size
    \item best ask price
    \item best ask size
    \item spread
    \item mid-price
    \item level-1 imbalance
    \item second bid price
    \item second bid size
    \item second ask price
    \item second ask size
    \item event-side feature
\end{enumerate}

The remaining 4 are one-step history-delta features appended by
\texttt{DataLoader.augment\_feature\_vector}:
\begin{enumerate}
    \setcounter{enumi}{12}
    \item one-step best bid delta
    \item one-step best ask delta
    \item one-step mid-price delta
    \item one-step imbalance delta
\end{enumerate}

The label policy is \textbf{horizon-based}. For each example, we compare
the current midpoint to the midpoint 50 events later:
\begin{itemize}
    \item if the future midpoint is higher by more than the threshold, the
    label is \texttt{buy}
    \item if it is lower by more than the threshold, the label is
    \texttt{sell}
    \item otherwise the label is \texttt{hold}
\end{itemize}

This pipeline is implemented in \texttt{DataLoader.build\_training\_dataset} and
related helper methods. The final training flow in \texttt{training.py}
computes train/validation splits, standardizes features with training-set mean
and standard deviation, applies class-weighted cross-entropy, trains a PyTorch
feed-forward network, and saves the latest training artifacts:
\begin{itemize}
    \item \texttt{latest\_model.pt}
    \item \texttt{latest\_training\_metrics.json}
    \item \texttt{latest\_training\_data.csv}
\end{itemize}

The shipped baseline saved-state artifacts remain \texttt{model.pt} and
\texttt{training\_metrics.json}, so the TA can always return to a known working
checkpoint.

\subsection*{3.4 User Interactivity}
The TA-facing interface is a browser UI launched by \texttt{main.py} and
served by \texttt{web.py}. Running \texttt{main.py} in PyCharm, or running
\texttt{python3 main.py}, starts the browser workflow at
\texttt{http://127.0.0.1:9000}. The browser UI has four tabs:
\begin{itemize}
    \item \texttt{Simulate}
    \item \texttt{Train}
    \item \texttt{Artifacts}
    \item \texttt{Charts}
\end{itemize}

Those tabs expose the same underlying simulator and training pipeline from
different angles. \texttt{Simulate} chooses a synthetic scenario or CSV replay,
controls replay speed, and selects the overlay mode. \texttt{Train} runs the
packaged internal datasets, the packaged LOBSTER message-file path, or a custom
CSV path. \texttt{Artifacts} reports the active overlay state, saved
checkpoints, metrics files, and latest training outputs. \texttt{Charts}
renders order-book depth and execution-price history after a simulation. The
backup terminal path uses \texttt{terminal\_menu.py}, but the intended TA
workflow is the browser UI.

The active overlay choice is stored in \texttt{active\_model.json}, which
makes the browser workflow persistent across runs. For submission readiness, we
split the orchestration code into \texttt{config.py}, \texttt{simulation.py},
\texttt{training.py}, \texttt{api\_payloads.py}, \texttt{web.py}, and
\texttt{terminal\_menu.py}, while keeping \texttt{main.py} as a small entry
point.

\subsection*{3.5 Overall Computational Complexity}
Quantyze demonstrates computational complexity in all three major ways
highlighted by the project rubric.

\begin{itemize}
    \item \textbf{Non-trivial dataset processing and model training:} the system
    supports multiple input formats, infers paired LOBSTER order-book files,
    separates replay-compatible and non-replayable events, builds a shared
    training dataset, normalizes features, computes class weights, and saves
    checkpoints and metric artifacts.
    \item \textbf{Non-trivial algorithms implemented by us:} the BST-based order
    book, FIFO price-level queues, price-time-priority matching engine, and
    cancellation/indexing logic are all custom project code rather than thin
    wrappers around third-party libraries.
    \item \textbf{Non-trivial reporting and interactivity:} the browser UI,
    terminal menu, saved-state workflow, metrics inspection, and
    execution-log reporting make the program substantially more interactive
    than a single fixed command.
\end{itemize}

\section*{4. Instructions for Obtaining Datasets and Running the Program}
The TA-facing run path is a single browser-first workflow. The most important
distinction is that \textbf{simulation source} and \textbf{model source} are
separate:
\begin{itemize}
    \item the \textbf{simulation source} is the event stream being replayed,
    such as \texttt{balanced}, \texttt{low\_liquidity},
    \texttt{high\_volatility}, or a custom CSV
    \item the \textbf{model source} is the classifier overlay currently
    selected for inference, such as the shipped \texttt{model.pt} baseline or
    a newly trained \texttt{latest\_model.pt}
\end{itemize}

Keeping them separate lets the TA test the same classifier overlay under
different market conditions and retrain a model without overwriting the shipped
baseline artifacts. We found this distinction important because the simulator
is the core system, while the model is an optional layer on top of it.

\subsection*{4.1 Installing Libraries}
Install the libraries listed in \texttt{requirements.txt} before running the
program.

\subsection*{4.2 Obtaining Datasets}
The TA downloads \texttt{quantyze\_datasets.zip} from our final MarkUs
submission and extracts it beside \texttt{main.py}. After extraction, the
project should see:
\begin{itemize}
    \item \texttt{sample\_internal.csv}
    \item \texttt{huge\_internal.csv}
    \item \texttt{aapl\_lobster\_2012-06-21\_message\_5level\_sample.csv}
    \item \texttt{aapl\_lobster\_2012-06-21\_orderbook\_5level\_sample.csv}
    \item \texttt{instructions.txt}
    \item \texttt{model.pt}
    \item \texttt{training\_metrics.json}
\end{itemize}

The shortest TA path should start with \texttt{sample\_internal.csv}. The
larger internal dataset and the packaged LOBSTER pair are both slower-loading
options for a more thorough optional check.

As the TA trains or changes overlay state, the runtime may also create or
update:
\begin{itemize}
    \item \texttt{latest\_model.pt}
    \item \texttt{latest\_training\_metrics.json}
    \item \texttt{latest\_training\_data.csv}
    \item \texttt{active\_model.json}
    \item \texttt{log.json}
\end{itemize}

\subsection*{4.3 Workflow Overview}
The primary grading path is:
\begin{enumerate}
    \item extract \texttt{quantyze\_datasets.zip} beside \texttt{main.py}
    \item open the project in PyCharm
    \item run \texttt{main.py}
    \item open \texttt{http://127.0.0.1:9000}
    \item use the \texttt{Simulate}, \texttt{Train}, \texttt{Artifacts}, and
    \texttt{Charts} tabs
\end{enumerate}

This path makes four things explicit at all times:
\begin{itemize}
    \item what event source is being replayed
    \item which checkpoint, if any, is active
    \item where that checkpoint came from
    \item what to do next: simulate, train, or inspect artifacts
\end{itemize}

The packaged datasets and baseline artifacts are local, but the browser
\texttt{Charts} tab loads \texttt{Chart.js} from a standard CDN. That means
browser chart rendering assumes a normal internet connection even though the
submission data and checkpoint files are bundled locally.

\subsection*{4.4 Recommended TA Demo}
The clearest way to understand the project is the following walkthrough after
running \texttt{main.py} and opening the printed local URL:
\begin{enumerate}
    \item Start on the \texttt{Simulate} tab and run the default
    \texttt{balanced} scenario. This is the fastest simulator-first path and
    produces the same summary metrics described later in this section.
    \item On the same tab, optionally run \texttt{low\_liquidity} and
    \texttt{high\_volatility}. These scenarios show that the same overlay can
    be evaluated under different event streams without changing the core
    matching engine.
    \item Open \texttt{Train}. For a short demonstration, use
    \texttt{sample\_internal.csv}. For a baseline-scale demonstration, use
    \texttt{huge\_internal.csv}. \\
    For the real-data path, use
    \texttt{aapl\_lobster\_2012-06-21\_message\_5level\_sample.csv}. The
    latter two are both much larger and take longer to load. After training,
    the UI writes the latest output files and exposes the new checkpoint in the
    saved artifact status.
    \item Open \texttt{Artifacts}. Check the overlay status, saved files, and
    baseline versus latest metrics.
    \item Open \texttt{Charts} after a simulation. This shows order-book depth
    and execution-price history based on the replay that just ran.
    \item Re-run \texttt{balanced} under a different overlay to isolate the
    difference between \textit{market replay} and \textit{agent overlay}
    behavior.
\end{enumerate}

\subsection*{4.5 How the TA Should Interpret the Output}
The browser UI summaries and terminal menu contain the same two classes of
information, and keeping them separate is the key to understanding Quantyze.

\subsubsection*{Simulation Run Configuration}
Before each run, the program prints:
\begin{itemize}
    \item the \textbf{event source}, such as \texttt{synthetic balanced}
    or a custom replay path
    \item the \textbf{simulation overlay mode}, such as \texttt{baseline},
    \texttt{latest}, or \texttt{none}
    \item the \textbf{simulation overlay path}
    \item the \textbf{overlay provenance}, such as the packaged baseline
    checkpoint or a newly trained dataset like \texttt{sample\_internal.csv}
\end{itemize}

This header tells the TA which market is being replayed and which classifier
overlay, if any, is observing it. They are related, but they are not the same
thing.

\subsubsection*{Matching-Engine Metrics}
The following summary values describe the replay itself:
\begin{itemize}
    \item \texttt{Total Filled}: total executed quantity produced by the
    matching engine
    \item \texttt{Fill Count}: number of executions
    \item \texttt{Cancel Count}: number of cancellations processed
    \item \texttt{Average Slippage}: average execution-price deviation under the
    current slippage calculation
    \item \texttt{Spread}: final bid-ask spread
    \item \texttt{Mid Price}: final midpoint
\end{itemize}

These values are properties of the event stream and the matching engine. If the
TA replays the same scenario with \texttt{baseline}, \texttt{latest}, and
\texttt{none}, the market metrics should stay the same because the underlying
event stream is unchanged.

\subsubsection*{Agent Overlay Metric}
If an active checkpoint is loaded, the program also prints
\texttt{Agent Overlay Mark-to-Market P\&L}. This value needs to be interpreted
carefully.

It is \textbf{not} a claim that the model is a profitable trading strategy. In
the current system, the neural network is trained from the midpoint move
50 events ahead and maps each step to \texttt{buy}, \texttt{sell}, or
\texttt{hold}. During
simulation, that prediction is converted into a simple paper-trading overlay
that increments or decrements inventory by one unit whenever the simulator
observes fills. As a result:
\begin{itemize}
    \item a strong classification accuracy does not automatically imply a strong
    trading P\&L
    \item a negative mark-to-market value can reflect inventory accumulation in
    a thin market rather than a broken simulator
    \item the ML component should be interpreted as a proof-of-concept
    prediction overlay, not as a finished trading strategy
\end{itemize}

For example, \texttt{low\_liquidity} often ends with a wider spread and fewer
fills than \texttt{balanced}. A model can therefore finish with non-zero
inventory and a negative mark-to-market value even when the matching engine
itself is behaving correctly.

\subsubsection*{Training Metrics}
The \texttt{Train} tab and the terminal menu report validation accuracy,
majority-class baseline, per-class recall, and the confusion matrix. These
values measure how
well the classifier reproduces the label rule derived from future midpoint
movement. They do not directly measure whether the simulated agent will earn
positive P\&L on every synthetic scenario.

\subsection*{4.6 Optional Raw LOBSTER Check}
If the TA wants to test the raw LOBSTER path, they do \textbf{not} need to
download external files. They can use the packaged training-menu option for
\texttt{aapl\_lobster\_2012-06-21\_message\_5level\_sample.csv}. The paired
order-book path is inferred automatically from that message filename, which
maps to
\texttt{aapl\_lobster\_2012-06-21\_orderbook\_5level\_sample.csv}.

The same loader logic also works for a custom raw LOBSTER message CSV path: it
infers the paired order-book path by replacing \texttt{\_message\_} with
\texttt{\_orderbook\_}.

\section*{5. Changes from Proposal to Final Submission}
The final project is cleaner and narrower than the proposal. The simulator
remained the core deliverable, and the ML component stayed as an optional
overlay rather than becoming the center of the project.

On the implementation side, we ended with a clearer module split than we
originally imagined: \texttt{config.py} holds shared constants and overlay
state, \texttt{simulation.py} owns replay orchestration, \texttt{training.py}
owns classifier training, \texttt{api\_payloads.py} owns browser-API response
serialization, \texttt{web.py} serves the browser UI,
\texttt{terminal\_menu.py} owns the backup terminal flow, and \texttt{main.py}
is now a small entry point.

Our frontend plan also changed. The proposal left room for a heavier
React/Plotly-style interface, but the final submission uses a lightweight
Flask/browser UI instead. That choice kept the focus on the tree-based order
book, matching engine, and training pipeline while still giving the TA a clear
PyCharm-friendly workflow.

The data plan became more concrete as well. Instead of treating external data
abstractly, the final submission cleanly separates synthetic replay scenarios,
packaged internal CSVs for fast retraining, and one packaged raw LOBSTER pair
for the real-data preprocessing path.

\section*{6. Discussion}
Our goal was to model the mechanics of a limit order book using trees and to
connect that simulator to a simple market-action classifier. The final project
does meet that goal, though the machine learning side is more modest than the
simulator itself.

On the simulation side, the output is meaningful and easy to interpret. In the
current default \texttt{balanced} run, the simulator reports 1748.00 total
filled volume, 374 fills, 128 cancellations, a final spread of 0.60, and a
final midpoint of 100.00. Those values show that the event sequence
is not just inserting passive orders forever; it is generating executions,
consuming liquidity, and rebuilding the book over time. That mattered to us
from the beginning, because one of the main motivations for the project was to
show the difference between a quoted market price and the mechanics required to
fill an order.

We also think the project succeeds as a data-structures project. The binary
search trees are not decorative. They are the central representation of the
market. Maintaining sorted price levels, extracting the best bid and best ask,
removing empty levels, and preserving FIFO priority within a level all fit the
tree-based design naturally. That gives the project real computational
substance beyond simple data loading or basic statistics.

The machine learning component works, but it needs to be interpreted carefully.
The packaged baseline checkpoint and its accompanying metrics artifact now show
a validation accuracy of approximately 0.81644 compared to a majority-class
baseline of approximately 0.66184 on \texttt{huge\_internal.csv}. These values
are useful for confirming that the training pipeline, checkpoint loader, and
saved-state workflow function correctly, but they are still tied to a packaged
internal dataset rather than a difficult real-market benchmark.

The run-time P\&L shown in the simulator should also be read as an agent
overlay metric rather than as proof of profitability. The neural network is
trained on future midpoint labels, not on a direct trading reward, and the
simulation overlay uses those labels to build a simple paper position over
time. Because of that, the classifier can behave as designed while still
producing a negative mark-to-market value in a thin or volatile synthetic
scenario.

When we tested on public LOBSTER sample data during development, the task was
much harder and performance was only modestly above a naive baseline. For that
reason, we do not present the current model as a strong predictive system for
real markets. In this project, its role is better understood as a small
proof of concept that connects book-state features to model output inside a
complete simulator.

That still answers the project goal in a meaningful way. The goal was never
just to maximize predictive accuracy. It was to build a full computational
pipeline from market events to order-book state to derived features to model
output. In that sense, the ML side is successful: it uses both synthetic and
real-data paths, produces saved checkpoints, supports inference inside the
simulator, and can be retrained from the browser UI or the terminal menu.

We also ran into clear limitations. Synthetic data is easy to test and reason
about, but it is not enough to justify a strong predictive claim. Real LOBSTER
data is much richer, but it brings preprocessing issues, class-balance issues,
and a harder learning problem. The model itself is also simple: we used a small
feed-forward network over handcrafted features rather than a stronger temporal
model. That was a reasonable CSC111 choice, but it limits what the model can
learn about market dynamics. Finally, our presentation scope narrowed over the
course of the project. The proposal imagined a more polished UI layer, but most
of our time went into the simulation engine, dataset handling, and ML pipeline.

If we were to continue the project, the next steps would be to improve the
visualization layer, add a stronger sequence-aware model, and evaluate the
trained agent with more realistic trading metrics across multiple market days.
Even with those limitations, we think the current submission succeeds as a
CSC111 project because it combines tree-based design, non-trivial event
processing, model training, and interpretable output in one coherent system.
More importantly, it still reflects the original motivation behind the project:
to make the hidden mechanics of a market feel concrete rather than abstract.

\section*{7. References}
\begin{enumerate}
    \item LOBSTER. (n.d.). \textit{Sample data}. Retrieved March 31, 2026, from
    \url{https://data.lobsterdata.com/info/DataSamples.php}
    \item PyTorch. (n.d.). \textit{PyTorch}. Retrieved March 31, 2026, from
    \url{https://pytorch.org/}
    \item Investopedia. (n.d.). \textit{Limit order book}. Retrieved March 31,
    2026, from
    \url{https://www.investopedia.com/terms/l/limitorderbook.asp}
    \item Investopedia. (n.d.). \textit{Depth of market (DOM)}. Retrieved March
    31, 2026, from
    \url{https://www.investopedia.com/terms/d/depth-of-market.asp}
    \item QuestDB. (n.d.). \textit{Limit order book}. Retrieved March 31, 2026,
    from \url{https://questdb.com/glossary/limit-order-book/}
    \item Gould, M.\,D., Porter, M.\,A., Williams, S., McDonald, M., Fenn,
    D.\,J., \& Howison, S.\,D. (2013). Limit order books.
    \textit{Quantitative Finance}, 13(11), 1709--1742.
    Retrieved March 31, 2026, from \url{https://arxiv.org/abs/1012.0349}
\end{enumerate}

\end{document}
